---
title: "RPM RPR2 Allo notebook"
author: "Abbie Ireland, Darren Tyson"
date: "2025-06-24"
output: html_notebook
---

Modified from original code in [Fig2l-o_ExtFig4h-j_RPM_RPR2_Allografts_Final.R](Preprocessing_for_reference/R_Code/Fig2l-o_ExtFig4h-j_RPM_RPR2_Allografts_Final.R).

## Related to Extended Data Fig 4i-n
```{r}
suppressPackageStartupMessages({ 
    library(tidyverse)
    library(SingleCellExperiment)
    library(ggplot2)
    library(cowplot)
    library(Seurat)
    library(SeuratObject)
    library(viridis)
    library(ggpubr)
})
```

### Load the data
```{r}
RPMvRPR2 <- readRDS("../data/05_2025_RPMvRPR2_Allografts_Final.rds")
```

#### Define colors
```{r}
unid_cols <- c("turquoise3","maroon")

pheno_col <- c("brown2","darkorchid4","dodgerblue","#66A61E","orange","turquoise4","maroon","gray90")

colors <- c('#F39B7F99','brown1','#ff7f0e', 'turquoise', '#2ca02c', 'deepskyblue4', '#e7298a',  
            '#4ba4d3',  '#A0522D', '#a1c299', 'gold', '#984ea3', '#3C548899', '#377eb8',  '#d62728',
            'navyblue', '#a1c299', 'purple','gray','maroon','gray20')

```


#### Ext Data Fig 4i
```{r fig.height=4, fig.width=4}
DimPlot(RPMvRPR2, group.by='Genotype', cols=unid_cols, reduction='umap', shuffle=TRUE, label=FALSE) + NoAxes()
```

#### Ext Data Fig 4j
```{r fig.height=4, fig.width=4}
# Define color vector and use to plot UMAP in Fig. 2m #
DimPlot(RPMvRPR2, group.by='leiden_scVI_1.2', cols=colors, reduction='umap', pt.size = 0.25,
        label=TRUE, label.size=5) & NoAxes() & NoLegend() & ggtitle("Ext Data Fig 4j clusters")

```

#### Ext Data Fig. 4l
UMAP
```{r fig.height=4, fig.width=5.5}
DimPlot(RPMvRPR2, group.by='RPM_Pheno', cols=pheno_col, reduction='umap',
        shuffle=TRUE) + NoAxes() + 
    ggtitle("RPM Cell State")

```

#### Function for plotting expression in violin plots by genotype
```{r}
plotVlnFeatureByGeno <- function(feature, seurat_obj=RPMvRPR2, y_max=NA, tit=NA, alpha=0.5) {
    if(is.na(tit)) tit <- feature
    a <- VlnPlot(seurat_obj, features = c(feature), 
                 group.by=c("Genotype"), same.y.lims=FALSE, 
                 cols=c("turquoise3","maroon"), pt.size=.025,
                 alpha=alpha, ncol=1, y.max=y_max)
    # Add mean point and do wilcoxon rank sum test 
    a + stat_summary(fun = mean,
                     geom = "point",
                     shape = 18,
                     size = 2,
                     color = "yellow",
                     position = position_dodge(width = 0.9)) + 
        ggtitle(tit) + 
        theme(axis.title.x = element_blank()) + 
        stat_compare_means(method = "wilcox.test", 
                           label = "p.signif",
                           hide.ns = TRUE,
                           comparisons = list(c("RPM", "RPR2"))) +
        theme(legend.position = "none")
}
```

#### Extended Data Fig. 4i
```{r fig.height=3, fig.width=2, message=FALSE, warning=FALSE}
plotVlnFeatureByGeno("Ascl1")
plotVlnFeatureByGeno("Neurod1")
plotVlnFeatureByGeno("Atoh1")
plotVlnFeatureByGeno("Pou2f3")
plotVlnFeatureByGeno("Mycl")
plotVlnFeatureByGeno("Myc")
```

#### Ext Data Fig 4k
```{r fig.height=3, fig.width=3.5, message=FALSE, warning=FALSE}
FeaturePlot(RPMvRPR2, features = c("NE_spearman"), pt.size=0.2,
            reduction='umap', order=TRUE) + 
    scale_color_viridis(option="rocket",direction=-1) & NoAxes()

```


#### Ext Data Fig 4m
```{r fig.height=3, fig.width=2, message=FALSE, warning=FALSE}
plotVlnFeatureByGeno("A_Archetype1", alpha=0.1)
plotVlnFeatureByGeno("A2_Archetype1", alpha=0.1)
plotVlnFeatureByGeno("N_Archetype1", alpha=0.1)
plotVlnFeatureByGeno("P_Archetype1", alpha=0.1)
```


#### Ext Data Fig 4n
```{r fig.height=3, fig.width=2, message=FALSE, warning=FALSE}
plotVlnFeatureByGeno("ASCL1_Targets1", alpha=0.1)
plotVlnFeatureByGeno("NEUROD1_Targets1", alpha=0.1)
plotVlnFeatureByGeno("POU2F3_Targets1", alpha=0.1)
plotVlnFeatureByGeno("MYC_Targets1", alpha=0.1)
```


```{r fig.height=3, fig.width=1.75, message=FALSE, warning=FALSE}
x <- VlnPlot(RPMvRPR2,features = c("NE_spearman"),
             same.y.lims=FALSE, group.by=c("Geno_C8"),
             cols=c("turquoise3","maroon","#A0522D"), alpha=.2, ncol=1)
x + geom_boxplot(fill=c("turquoise3","maroon","#A0522D"), 
                 alpha=1, position = position_dodge(width = .2),
                 size=0.5, color="black", notch=TRUE, notchwidth=0.3, 
                 outlier.shape = 2, outlier.colour=NA) + 
    ggtitle("NE score") + NoLegend() + labs(x = NULL, y = "NE score")

```

