---
title: "RPM RPMA CellTag In Vitro Notebook"
output: html_notebook
date: 2025-07-10
author: "Abbie Ireland and Darren Tyson"
---

### Modified original file [ExtFig7_RPM_RPMA_CellTag_InVitrotoInVivo_Final.R](Preprocessing_for_reference/R_Code/ExtFig7_RPM_RPMA_CellTag_InVitrotoInVivo_Final.R)  

### Related to
* Extended Data Fig 7b,e,f


```{r}
suppressPackageStartupMessages({ 
    library(tidyverse)
    library(SingleCellExperiment)
    library(ggplot2)
    library(cowplot)
    library(Seurat)
    library(ggpubr)
    library(ggnewscale)
})
```


### Load data
```{r}
TBO_seurat_invitro <- readRDS("../data/05.15.25_RPM_RPMA_WT_Organoids_wCT_Clones.rds")
clones <- readRDS("../data/05.18.25_RPM_RPMA_invitro_CTclones_all.rds")
clones_subset <- readRDS("../data/05.18.25_RPM_RPMA_invitro_CTclones_matchinginvivo.rds")
```
#### Define colors for leiden clusters
```{r}
colors_leiden <- c('deepskyblue4', '#ff7f0e','#2ca02c', 'brown1', 'purple', '#A0522D', 
                   'pink2',  '#a1c299',  '#4ba4d3', '#b5d7e4','#ef9171','#a6eb9a')

pheno_cols <- c('#1B9E77', '#D95F02', '#7570B3', '#E7298A', '#66A61E', '#E6AB02', '#A6761D', '#666666')[1:5]
names(pheno_cols) <- levels(clones_subset@meta.data$CellType)

pattern_colors <- c("Pattern_1" = "orange", "Pattern_2" = "green", "Pattern_5"="purple","Pattern_4"="blue")

```

#### Extended Data Fig 7b
```{r fig.height=3, fig.width=4}
DimPlot(TBO_seurat_invitro, group.by='leiden_scVI_1.2', 
        cols=colors_leiden, reduction='umap', label=TRUE,
        label.size=8) & NoAxes()

```

```{r fig.height=3, fig.width=6}
DimPlot(TBO_seurat_invitro, group.by='CellType_Sig_Based', cols="Dark2", reduction='umap', 
        label=FALSE, label.size=6) & NoAxes() 
```


```{r}
# Subset clones to just those with matching in vivo clones
invivo_clones <- read.csv("../data/cloneswmatch.csv")
clone_match <- invivo_clones$Clone
Idents(clones)<-'CellTag_Clone'
clones_subset <- subset(clones, idents=clone_match)
table(clones_subset$CellTag_Pattern)
```


```{r}
Idents(TBO_seurat_invitro) <- 'CellTag_Binary'

```

```{r}
Idents(clones_subset) <- 'CellType_Sig_Based'

x <- table(clones_subset@meta.data$CellTag_Clone,Idents(clones_subset))
proportions <- as.data.frame(100*prop.table(x, margin = 1))

colnames(proportions)<-c("Cluster", "Sample", "Frequency")
proportions$Cluster<-factor(proportions$Cluster, clone_match)
```

#### Ext. Data Fig. 7e
```{r message=FALSE, warning=FALSE}
# Add pattern annotation 
x <- invivo_clones$Pattern

# Create annotation data
annotation_df <- data.frame(
  Sample = levels(proportions$Cluster),  # match x-axis labels
  Group = x,  # Define annotation groups
  y = -0.5  # Place below x-axis, adjust as needed
)

p <- ggplot(proportions, aes(fill=Sample, y=Frequency, x=Cluster)) + 
    geom_bar(position="stack", stat="identity")

p + scale_fill_manual(values=pheno_cols) +  
    theme_bw() + theme(axis.text.y = element_text(size=20), 
                       axis.text.x=element_text(size=14), axis.title.x =element_text(size=14), 
                       axis.title.y = element_text(size=18), legend.text = element_text(size=12), 
                       legend.title = element_text(size=18))+rotate_x_text(size=7,angle = 90) +
    scale_fill_manual(values = pheno_cols, name="Basal cell state") +
    theme_bw() +
    theme(
        axis.text.y = element_text(size = 20),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 18)
    ) +
    rotate_x_text(size = 7, angle = 90) +
  
  # Reset fill scale so annotation can have its own
    new_scale_fill() +
  
  # Add annotation bar
    geom_tile(data = annotation_df,
              aes(x = Sample, y = y, fill = Group),
              height = 3,
              inherit.aes = FALSE
    ) +
  scale_fill_manual(name = "In vivo clonal pattern", values = pattern_colors,
                    guide = guide_legend(override.aes = list(size = 5)))
```

#### Extended Data Fig 7f
Output from these plots were manually overlaid onto full UMAP with all cells in grey using Adobe Illustrator
```{r}
#Plot RPM patterns 1, 2, and 5 that have matching in vivo clones
Idents(clones_subset) <- 'CellTag_Pattern'

plotCloneDyn <- function(pat) {
    ss <- subset(clones_subset, idents=pat)
    DimPlot(ss, group.by="CellType_Sig_Based", reduction='umap', order=TRUE, 
            cols=pheno_cols, pt.size=6, shuffle=TRUE) + ggtitle(pat) & 
        NoAxes() & 
        theme(legend.background = element_rect(fill = "transparent"),
              legend.box.background = element_rect(fill = "transparent"),
              panel.background = element_rect(fill = "transparent"),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              plot.background = element_rect(fill = "transparent", color = NA))
    
}
```

```{r}
lapply(c("Pattern_1","Pattern_2","Pattern_5"), plotCloneDyn)

```

