---
title: "Extended Data RPM K5 vs CGRP notebook"
output: html_notebook
date: "2025-06-17"
author: "Abbie Ireland, Darren Tyson"
---

```{r}
OVERWRITE <- FALSE
```

### Load the necessary libraries
```{r}
suppressPackageStartupMessages({ 
    library(tidyverse)
    library(ggplot2)
    library(cowplot)
    library(Seurat)
    library(viridis)
    library(ggpubr)
    library(dplyr)
    library(FSA)       # For dunnTest
    library(rstatix)   # Optional, nice formatting of results
    library(patchwork)
    library(gridExtra)
})
```
###  P star function
```{r}
# Function to assign p-value stars
get_p_stars <- function(p) {
    if (p < 0.0001) return("****")
    else if(p < 0.001) return("***")
    else if (p < 0.01) return("**")
    else if (p < 0.05) return("*")
    else return("ns")
}
```
### Load the data
```{r}
# RPM_K5vCGRP <- readRDS("../data/05_2025_RPMK5vCGRP_Norm_Allsigs_New.rds") # original AI
RPM_K5vCGRP <- readRDS("../data/RPM_K5vCGRP_Seurat.Rds") # From BasalDataCleanup_20250612.ipynb
```

```{r}
unid_cols <- c("darkorchid4","orange")
```


### Plot UMAP of samples
#### Ext Data Fig 2g
```{r fig.height=3, fig.width=4}
DimPlot(RPM_K5vCGRP, group.by='Cre', cols=unid_cols, reduction='umap', shuffle=TRUE) + NoAxes()
```
#### Ext Data Fig 2g
```{r fig.height=3, fig.width=6}
DimPlot(RPM_K5vCGRP, split.by='Cre', group.by='Cre', cols=unid_cols, reduction='umap', shuffle=TRUE) + NoAxes()
```


### Leiden clustering
#### Ext Data Fig 2i
```{r fig.height=3, fig.width=4}
# Plots to generate Fig 1i
# UMAP by Leiden cluster
colors<-c('#F39B7F99','brown1','#ff7f0e', 'turquoise', '#2ca02c', 'deepskyblue4', '#e7298a',  '#0aa6d8',  '#A0522D', '#a1c299', 'gold', '#984ea3', '#3C548899', '#377eb8',  '#d62728','navyblue', '#a1c299', 'purple','gray','maroon','gray20')

DimPlot(RPM_K5vCGRP,group.by='leiden_scVI_1.3', cols=colors, reduction='umap', label=TRUE, label.size=6) + NoAxes()
```

### What % of cells occupy each cluster for each sample?
```{r}
Idents(RPM_K5vCGRP)<-'Cre'

x <- table(Idents(RPM_K5vCGRP),RPM_K5vCGRP@meta.data$leiden_scVI_1.3)
proportions <- as.data.frame(100*prop.table(x, margin = 1))

colnames(proportions) <- c("Cluster", "Sample", "Frequency")
```


#### Stacked bar charts
```{r fig.height=5, fig.width=4}
p <- ggplot(proportions, aes(fill=Sample, y=Frequency, x=Cluster)) + 
    geom_bar(position="stack", stat="identity")

p + scale_fill_manual(values=colors) + theme_bw() + 
    theme(axis.text.y = element_text(size=20), 
          axis.text.x = element_text(size=20, angle=45, vjust=1, hjust=1), 
          axis.title.x = element_text(size=20), 
          axis.title.y = element_text(size=20), 
          legend.text = element_text(size=20), 
          legend.title = element_blank()) + 
    labs( x = NULL, y = "Frequency (%)")

```

### ANP expr by Cre
#### Ext Data Fig. 2h
```{r message=FALSE, warning=FALSE}
plots <- VlnPlot(
    RPM_K5vCGRP, 
    features = c("Ascl1", "Neurod1", "Pou2f3"), 
    same.y.lims = FALSE, 
    group.by = "Genotype", 
    split.by = "Cre", 
    split.plot = TRUE, 
    cols = c("darkorchid4", "orange"), 
    pt.size = 0.01
)

# Apply theme to all plots
plots <- lapply(plots, function(p) {
    p + theme(axis.title.x = element_blank(),
              axis.text.x = element_text(angle = 45, hjust = 1))
})

# Combine with patchwork
wrap_plots(plots, ncol = 3)
```

### Function for Fig. 1j-m
#### Violin plot by Cre type
```{r}
plotVlnFeatureByCre <- function(feature, seurat_obj=RPM_K5vCGRP, y_max=NA, tit=NA) {
    if(is.na(tit)) tit <- feature
    a <- VlnPlot(seurat_obj, features = c(feature), 
                 group.by=c("Cre"), same.y.lims=FALSE, 
                 cols=c("darkorchid4","orange"), pt.size=.01,
                 alpha=0.05, ncol=1, y.max=y_max)
    # Add mean point and do wilcoxon rank sum test 
    a + stat_summary(fun = mean,
                     geom = "point",
                     shape = 18,
                     size = 2,
                     color = "yellow",
                     position = position_dodge(width = 0.9)) + 
        ggtitle(tit) + 
        theme(axis.title.x = element_blank()) + 
        stat_compare_means(method = "wilcox.test", 
                           label = "p.signif",
                           hide.ns = TRUE,
                           comparisons = list(c("CGRP", "K5"))) +
        theme(legend.position = "none")
}
```


#### Function to plot expression level by Leiden cluster
```{r}
plotVlnFeatureByLeiden <- function(feature, tit=NA){
    a <- VlnPlot(RPM_K5vCGRP, features = feature, 
                 group.by=c("leiden_scVI_1.3"), same.y.lims=FALSE, cols=colors, 
                 pt.size=.01, alpha=0.05, ncol=1)
    # Add mean point
    a + stat_summary(fun = mean,
                     geom = "point",
                     shape = 18,
                     size = 2,
                     color = "red",
                     position = position_dodge(width = 0.9)) + 
        ggtitle(tit) + 
        labs(x = "Leiden cluster", y = "Expression") +
        theme(legend.position = "none")
    
}

```



## Generate violin plot by Cre

### A, N, P ChIP target genes as scores
#### Ext. Data. Fig. 2j
```{r fig.height=4, fig.width=2}
plotVlnFeatureByCre(feature="ASCL1_Targets1", tit="ASCL1")
plotVlnFeatureByCre(feature="NEUROD1_Targets1", tit="NEUROD1")
plotVlnFeatureByCre(feature="POU2F3_Targets1", tit="POU2F3")

```


#### Ext Data Fig 2k
Neuroendocrine score (Spearman correlation coefficient) 
```{r fig.height=4, fig.width=3}
plotVlnFeatureByCre(feature="NE_spearman", tit="NE score")
```

#### Ext. Data. Fig. 2l
SCLC archetypes (Groves et al., 2022)
```{r fig.height=4, fig.width=2}
plotVlnFeatureByCre(feature="A_Archetype1", tit="A")
plotVlnFeatureByCre(feature="A2_Archetype1", tit="A2")
plotVlnFeatureByCre(feature="N_Archetype1", tit="N")
plotVlnFeatureByCre(feature="P_Archetype1", tit="P")
```

#### Statistics of variation of expression among Leiden clusters
```{r}
# Extract the data
expr_values <- FetchData(RPM_K5vCGRP, vars = c("A2_Archetype1", "leiden_scVI_1.3"))

# Rename columns for convenience
colnames(expr_values) <- c("expression", "group")

# Run Kruskal-Wallis test
kruskal_result <- kruskal.test(expression ~ group, data = expr_values)

# If significant, do Dunn's test (if set method to "none", that would be uncorrected)
if (kruskal_result$p.value < 0.05) {
  dunn_result <- dunnTest(expression ~ group, data = expr_values, method = "none")
  print(dunn_result)
}

# Convert to data frame if needed
dunn_df <- dunn_result$res

# Apply function to create a new column
dunn_df$p_stars <- sapply(dunn_df$P.adj, get_p_stars)

# View result, Use p-val stars to annotate plots
print(dunn_df)
```

#### Extended Data Fig. 2m
Human A N P scRNA seq sigs from Chan et al 
```{r fig.height=4, fig.width=2}
plotVlnFeatureByCre(feature="hSCLC_A1", tit="SCLC-A")
plotVlnFeatureByCre(feature="hSCLC_N1", tit="SCLC-N")
plotVlnFeatureByCre(feature="hSCLC_P1", tit="SCLC-P")
```

#### Ext Data Fig 2o
```{r fig.height=3, fig.width=4}
plotVlnFeatureByLeiden(feature="A2_Archetype1", tit="SCLC-A2 archetype")
plotVlnFeatureByLeiden(feature="Basal_hillock1", tit="Basal hillock")
plotVlnFeatureByLeiden(feature="Luminal_hillock1", tit="Luminal hillock")
```

### Perform kruskal-wallis test and post-hoc dunn's
Related to Fig. 2o
```{r}
# Extract the data
expr_values <- FetchData(RPM_K5vCGRP, vars = c("Luminal_hillock1", "leiden_scVI_1.3"))

# Rename columns for convenience
colnames(expr_values) <- c("expression", "group")

# Run Kruskal-Wallis test
kruskal_result <- kruskal.test(expression ~ group, data = expr_values)

# If significant, do Dunn's test (if set method to "none", that would be uncorrected)
if (kruskal_result$p.value < 0.05) {
  dunn_result <- dunnTest(expression ~ group, data = expr_values, method = "bonferroni")
}

# Convert to data frame if needed
dunn_df <- dunn_result$res

# Apply function to create a new column
dunn_df$p_stars <- sapply(dunn_df$P.adj, get_p_stars)
head(dunn_df)
```
#### Ext Data Fig 2n
```{r fig.height=6, fig.width=5, message=FALSE, warning=FALSE}
Idents(RPM_K5vCGRP) <- 'Cre'

x <- table(Idents(RPM_K5vCGRP),RPM_K5vCGRP@meta.data$leiden_scVI_1.3)
proportions <- as.data.frame(100*prop.table(t(x), margin = 1))

colnames(proportions) <- c("Cluster", "Sample", "Frequency")

ggplot(proportions, aes(fill=Sample, y=Frequency, x=Cluster)) +
    geom_bar(position="stack", stat="identity") + 
    scale_fill_manual(values=unid_cols) + theme_bw() + 
    theme(axis.text.y = element_text(size=20), 
          axis.text.x=element_text(size=20), 
          axis.title.x =element_text(size=20), 
          axis.title.y = element_text(size=20), 
          legend.text = element_text(size=20), 
          legend.title = element_text(size=20)) + NoLegend() +
    labs(x = "Leiden cluster", y = "% of cluster")

```


